Partitioned Local Depth (PaLD) Community Analyses in R

### 2023-05-02 r-journal/dagostino-mcgowan.Rmd
update with new

---
title: "Partitioned Local Depth (PaLD) Community Cluster Analyses in R"
author:
  - name: Lucy D'Agostino McGowan
    affiliation: Wake Forest University
    address:
    - Winston-Salem, NC
    - 27106
    email:  mcgowald@wfu.edu
  - name: Katherine Moore
    affiliation: Wake Forest Unversity
    address:
    - Winston-Salem, NC
    - 27106
    email: mooreke@wfu.edu 
  - name: Kenneth Berenhaut 
    affiliation: Wake Forest University
    address:
    - Winston-Salem, NC
    - 27106
    email: berenhks@wfu.edu 
abstract: >
  Partitioned Local Depth (PaLD) is a framework for a holistic consideration of the community structure of distance-based data. This paper describes an R package, \CRANpkg{pald}, for calculating Partitioned Local Depth (PaLD) probabilities, implementing community clustering analyses, and creating data visualizations to display community structure. We describe how to use the package as well as walk through several examples. 
output: rticles::rjournal_article
---

@@

We present a new package, \CRANpkg{pald}, for calculating Partitioned Local Depth (PaLD) probabilities, implementing community clustering analyses, and creating data visualizations to display community structure. This paper will describe how to use the package as well as walk through several examples.

@@

The function above creates a `dist` object. If converted to a matrix, this will be a $n\times n$ distance matrix, where $n$ corresponds to the number of observations in the original data frame, in this example $n = 6$.

@@

The final category of function is that for data visualization. We can begin by visualizing the points in the data frame `df` (Figure \ref{fig:fig1}). When visualizing these points, it is important to have the aspect ratio of the x and y axes equal to 1 so as to not distort the distances. When using the \CRANpkg{ggplot2} package for this visualization, one can use the `coord_fixed(ratio = 1)` function. If using the `plot()` function included in the base library, one can use the `asp = 1` argument.

@@

We can pass the cohesion matrix to the `plot_community_graphs()` function to view the relationship between points (Figure \ref{fig:fig2}). Notice in this plot the layout does not match that of the original data frame as seen in Figure \ref{fig:fig1}. Since our original data is two dimensional, it may be reasonable to use this as the layout. Figure \ref{fig:fig3} will make this update as well as update some of the aesthetics, such as making the labels more readable.

@@

We will demonstrate the utility of the \CRANpkg{pald} package in three clustering examples.

@@

The `community_clusters()` function can be used to identify the community cluster corresponding to each tissue sample. Since the output is a data frame, we can summarize the clusters using commonly used data analysis techniques. For demonstration purposes, we will use the \CRANpkg{dplyr} package to summarize the contribution of clusters.

@@

### Clustering generated data

@@

Here, we will demonstrate how to use `exdata3`. These points were generated from bivariate normal distributions with varying means and variances. There are eight "true" clusters.

@@

Cohesion is particularly useful when considering data with varying local density, see discussion in \citet{berenhaut2022social}.  Note that the PaLD algorithm is able to detect the eight natural groups within the data without the use of any additional inputs (e.g., number of clusters) nor optimization criteria.  Despite the user input of the "correct" number of clusters (i.e., $k = 8$) both *k*-means and hierarchical clustering did not give the desired result.

@@

This paper introduces the \CRANpkg{pald} package, demonstrating its utility for providing parameter-free clustering which can easily be implemented for a variety of data sets.

>>>>

---
title: "Partitioned Local Depth (PaLD) Community Analyses in R"
author:
  - name: Lucy D'Agostino McGowan
    affiliation: Wake Forest University
    address:
    - Winston-Salem, NC
    - 27106
    email:  mcgowald@wfu.edu
  - name: Katherine Moore
    affiliation: Wake Forest Unversity
    address:
    - Winston-Salem, NC
    - 27106
    email: mooreke@wfu.edu 
  - name: Kenneth Berenhaut 
    affiliation: Wake Forest University
    address:
    - Winston-Salem, NC
    - 27106
    email: berenhks@wfu.edu 
abstract: >
  Partitioned Local Depth (PaLD) is a framework for a holistic consideration of the community structure of distance-based data. This paper describes an R package, \CRANpkg{pald}, for calculating Partitioned Local Depth (PaLD) probabilities, implementing community analyses, and creating data visualizations to display community structure. We describe how to use the package as well as walk through several examples. 
output: rticles::rjournal_article
---

@@

We present a new package, \CRANpkg{pald}, for calculating Partitioned Local Depth (PaLD) probabilities, implementing community analyses, and creating data visualizations to display community structure. This paper will describe how to use the package, walk through several examples, and compare the method to commonly used techniques in R. Together, these demonstrate both the novelty of the method and utility of the implementation in package described.

@@

The function above creates a `dist` object. If converted to a matrix, this will be an $n\times n$ distance matrix, where $n$ corresponds to the number of observations in the original data frame, in this example $n = 6$.

@@

The final category of function is that for data visualization. We can begin by visualizing the points in the data frame `df` (Figure \ref{fig:fig1}). When visualizing these points, it is important to have the aspect ratio of the x and y axes equal to 1 so as to not distort distances. When using the \CRANpkg{ggplot2} package for this visualization, one can use the `coord_fixed(ratio = 1)` function. If using the `plot()` function included in the base library, one can use the `asp = 1` argument.

@@

We can pass the cohesion matrix to the `plot_community_graphs()` function to view the relationship between points (Figure \ref{fig:fig2}). Notice in this plot the layout does not match that of the original data frame as seen in Figure \ref{fig:fig1}. Since our original data is two dimensional, it may be reasonable to use this as the layout. Figure \ref{fig:fig3} will make this update as well as update some of the aesthetics, such as employing more readable labels.

@@

We will demonstrate the utility of the \CRANpkg{pald} package in three illustrative examples.

@@

The `community_clusters()` function can be used to identify the community cluster corresponding to each tissue sample. Since the output is a data frame, we can summarize the result using commonly employed data analysis techniques. For demonstration purposes, we will use the \CRANpkg{dplyr} package to summarize the contribution of clusters.

@@

### Community analysis for generated data

@@

Here, we will demonstrate how to use `exdata3`. These points were generated from bivariate normal distributions with varying means and variances. There are eight "true" communities.

@@

Cohesion is particularly useful when considering data with varying local density, see discussion in \citet{berenhaut2022social}.  Note that the PaLD algorithm is able to detect the eight natural groups within the data without the use of any additional inputs (e.g., number of clusters) nor optimization criteria.  Despite the user input of the "correct" number of clusters (i.e., $k = 8$) both *k*-means and hierarchical clustering did not give the desired result. For further examples, discussion, and theoretical results, see \citet{berenhaut2022social}.

@@

This paper introduces the \CRANpkg{pald} package, demonstrating its utility for providing parameter-free clustering which can easily be implemented for a variety of data sets. We provide example code as well as compare the method to commonly used clustering techniques, *k*-means and hierarchical clustering.

### 2023-05-02 r-journal/dagostino-mcgowan.tex
update with new

% !TeX root = RJwrapper.tex
\title{Partitioned Local Depth (PaLD) Community Cluster Analyses in R}
\author{by Lucy D'Agostino McGowan, Katherine Moore, and Kenneth Berenhaut}

@@

\abstract{%
Partitioned Local Depth (PaLD) is a framework for a holistic
consideration of the community structure of distance-based data. This
paper describes an R package, \CRANpkg{pald}, for calculating
Partitioned Local Depth (PaLD) probabilities, implementing community
clustering analyses, and creating data visualizations to display
community structure. We describe how to use the package as well as walk
through several examples.
}

@@

We present a new package, \CRANpkg{pald}, for calculating Partitioned
Local Depth (PaLD) probabilities, implementing community clustering
analyses, and creating data visualizations to display community
structure. This paper will describe how to use the package as well as
walk through several examples.

@@

The function above creates a \texttt{dist} object. If converted to a
matrix, this will be a \(n\times n\) distance matrix, where \(n\)
corresponds to the number of observations in the original data frame, in
this example \(n = 6\).

@@

\begin{Schunk}
\begin{Sinput}
graphs <- community_graphs(cohesion)
graphs[["G_strong"]]
\end{Sinput}
\begin{Soutput}
#> IGRAPH 4190215 UNW- 6 3 -- 
#> + attr: name (v/c), weight (e/n)
#> + edges from 4190215 (vertex names):
#> [1] A--B A--E D--F
\end{Soutput}
\end{Schunk}

@@

The final category of function is that for data visualization. We can
begin by visualizing the points in the data frame \texttt{df} (Figure
\ref{fig:fig1}). When visualizing these points, it is important to have
the aspect ratio of the x and y axes equal to 1 so as to not distort the
distances. When using the \CRANpkg{ggplot2} package for this
visualization, one can use the \texttt{coord\_fixed(ratio\ =\ 1)}
function. If using the \texttt{plot()} function included in the base
library, one can use the \texttt{asp\ =\ 1} argument.

@@

We can pass the cohesion matrix to the
\texttt{plot\_community\_graphs()} function to view the relationship
between points (Figure \ref{fig:fig2}). Notice in this plot the layout
does not match that of the original data frame as seen in Figure
\ref{fig:fig1}. Since our original data is two dimensional, it may be
reasonable to use this as the layout. Figure \ref{fig:fig3} will make
this update as well as update some of the aesthetics, such as making the
labels more readable.

@@

The \texttt{community\_clusters()} function can be used to identify the
community cluster corresponding to each tissue sample. Since the output
is a data frame, we can summarize the clusters using commonly used data
analysis techniques. For demonstration purposes, we will use the
\CRANpkg{dplyr} package to summarize the contribution of clusters.

@@

\begin{Schunk}
\begin{Sinput}
french_neighbors <- igraph::neighbors(cognate_graph_strong, "French")
french_neighbors
\end{Sinput}
\begin{Soutput}
#> + 8/87 vertices, named, from 4db266d:
#> [1] Italian         Ladin           Provencal       Walloon        
#> [5] French_Creole_C French_Creole_D Spanish         Catalan
\end{Soutput}
\end{Schunk}

@@

\hypertarget{clustering-generated-data}{%
\subsubsection{Clustering generated
data}\label{clustering-generated-data}}

@@

Here, we will demonstrate how to use \texttt{exdata3}. These points were
generated from bivariate normal distributions with varying means and
variances. There are eight ``true'' clusters.

@@

Cohesion is particularly useful when considering data with varying local
density, see discussion in \citet{berenhaut2022social}. Note that the
PaLD algorithm is able to detect the eight natural groups within the
data without the use of any additional inputs (e.g., number of clusters)
nor optimization criteria. Despite the user input of the ``correct''
number of clusters (i.e., \(k = 8\)) both \emph{k}-means and
hierarchical clustering did not give the desired result.

@@

This paper introduces the \CRANpkg{pald} package, demonstrating its
utility for providing parameter-free clustering which can easily be
implemented for a variety of data sets.

>>>>

% !TeX root = RJwrapper.tex
\title{Partitioned Local Depth (PaLD) Community Analyses in R}
\author{by Lucy D'Agostino McGowan, Katherine Moore, and Kenneth Berenhaut}

@@

\abstract{%
Partitioned Local Depth (PaLD) is a framework for a holistic
consideration of the community structure of distance-based data. This
paper describes an R package, \CRANpkg{pald}, for calculating
Partitioned Local Depth (PaLD) probabilities, implementing community
analyses, and creating data visualizations to display community
structure. We describe how to use the package as well as walk through
several examples.
}

@@

We present a new package, \CRANpkg{pald}, for calculating Partitioned
Local Depth (PaLD) probabilities, implementing community analyses, and
creating data visualizations to display community structure. This paper
will describe how to use the package, walk through several examples, and
compare the method to commonly used clustering algorithms.

@@

The function above creates a \texttt{dist} object. If converted to a
matrix, this will be an \(n\times n\) distance matrix, where \(n\)
corresponds to the number of observations in the original data frame, in
this example \(n = 6\).

@@

\begin{Schunk}
\begin{Sinput}
graphs <- community_graphs(cohesion)
graphs[["G_strong"]]
\end{Sinput}
\begin{Soutput}
#> IGRAPH aa6899c UNW- 6 3 -- 
#> + attr: name (v/c), weight (e/n)
#> + edges from aa6899c (vertex names):
#> [1] A--B A--E D--F
\end{Soutput}
\end{Schunk}

@@

The final category of function is that for data visualization. We can
begin by visualizing the points in the data frame \texttt{df} (Figure
\ref{fig:fig1}). When visualizing these points, it is important to have
the aspect ratio of the x and y axes equal to 1 so as to not distort
distances. When using the \CRANpkg{ggplot2} package for this
visualization, one can use the \texttt{coord\_fixed(ratio\ =\ 1)}
function. If using the \texttt{plot()} function included in the base
library, one can use the \texttt{asp\ =\ 1} argument.

@@

We can pass the cohesion matrix to the
\texttt{plot\_community\_graphs()} function to view the relationship
between points (Figure \ref{fig:fig2}). Notice in this plot the layout
does not match that of the original data frame as seen in Figure
\ref{fig:fig1}. Since our original data is two dimensional, it may be
reasonable to use this as the layout. Figure \ref{fig:fig3} will make
this update as well as update some of the aesthetics, such as employing
more readable labels.

@@

The \texttt{community\_clusters()} function can be used to identify the
community cluster corresponding to each tissue sample. Since the output
is a data frame, we can summarize the result using commonly employed
data analysis techniques. For demonstration purposes, we will use the
\CRANpkg{dplyr} package to summarize the contribution of clusters.

@@

\begin{Schunk}
\begin{Sinput}
french_neighbors <- igraph::neighbors(cognate_graph_strong, "French")
french_neighbors
\end{Sinput}
\begin{Soutput}
#> + 8/87 vertices, named, from 7ab7e40:
#> [1] Italian         Ladin           Provencal       Walloon        
#> [5] French_Creole_C French_Creole_D Spanish         Catalan
\end{Soutput}
\end{Schunk}

@@

\hypertarget{community-analysis-for-generated-data}{%
\subsubsection{Community analysis for generated
data}\label{community-analysis-for-generated-data}}

@@

Here, we will demonstrate how to use \texttt{exdata3}. These points were
generated from bivariate normal distributions with varying means and
variances. There are eight ``true'' communities.

@@

Cohesion is particularly useful when considering data with varying local
density, see discussion in \citet{berenhaut2022social}. Note that the
PaLD algorithm is able to detect the eight natural groups within the
data without the use of any additional inputs (e.g., number of clusters)
nor optimization criteria. Despite the user input of the ``correct''
number of clusters (i.e., \(k = 8\)) both \emph{k}-means and
hierarchical clustering did not give the desired result. For further
examples, discussion, and theoretical results, see
\citet{berenhaut2022social}.

@@

This paper introduces the \CRANpkg{pald} package, demonstrating its
utility for providing parameter-free clustering which can easily be
implemented for a variety of data sets. We provide example code as well
as compare the method to commonly used clustering techniques,
\emph{k}-means and hierarchical clustering.
